---
title: "Untitled"
author: "Andrew S. Rosenberg"
date: "1/12/2017"
output: ioslides_presentation
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## What is a Simulation?
Simulations are a way to use your computer like a laboratory. Many statistical principles such as bias and consistency are based theoretically on the idea of repeated sampling. In other words, if we do things over and over again, what would the answer look like? 

## Central Limit Theorem in Action

-  Let's use simulation to show how the Central Limit Theorem works. 
- We remember what that means, right? 
- If you take a bunch of samples from a distribution and sum them up or take their average, then the distribution of those sums will be Gaussian.

## Example
- Everyone in the class has a quarter and is standing on the center line of a football field.
- We flip 16 times. If heads, take one step to the right. If tails, take one step to the left.
- Do all your flips and take all your steps, then stand still.
- We won't be able to say were a single person will end up, but we will know how the distribution of positions will look.

## Let's do it
We can even be agnostic about the size of a given step. Let's assume that a step is between 0 and 1 feet, but need not be the same. 

```{r steps}
inputPanel(
  selectInput("n_sims", label = "Number of sims:",
              choices = c(10, 30, 100, 1000, 5000), selected = 10),
  
  sliderInput("n_flips", label = "Class Size",
              min = 5, max = 40, value = 20, step = 1)
)

renderPlot({
  positions <- replicate(input$n_sims, sum(runif(input$n_flips, -1, 1)))
  qplot(positions, geom = "histogram") + theme_bw()
})
```

So this is the final position of one person. We calculated all the steps and added them up to get the final position. Now, let's simulate this same procedure for 20 people.

```{r}
library(ggplot2)
positions <- replicate(20, sum(runif(16, -1, 1)))
qplot(positions, geom = "histogram") + theme_bw()
```

That's only one person, so let's crank up the number of simulations a bit.

```{r}
library(ggplot2)
positions <- replicate(1000, sum(runif(16, -1, 1)))
qplot(positions, geom = "histogram") + theme_bw()
```

This is pretty cool. Let's do one more example. I am going to sample from the chisq distribution 100 times, add them together, and plot the histogram.

```{r}
hist(rchisq(10000, df = 5))
hist(replicate(1000, sum(rchisq(10, df = 5))))
```


## Birthday Problem
As we have seen, we can use simulation to find out the probability that any two people in a room of 24 have the same birthday. However, I'm going to show you a slick way of doing this. In addition, I'll figure out how this probability changes with the size of the room.
```{r}
library(ggplot2)
library(data.table)
birthday_problem <- function(npeople,  
  seed = sample.int(.Machine$integer.max, 1))
{
  days <- seq(1, 365, 1)
  room <- sample(days, npeople, replace = TRUE)
  ifelse(length(unique(room)) < npeople, 1, 0)
}
#birthday_problem(npeople = 24)

simulate_birthday_problem <- function(nsims, npeople)
{
  sum(replicate(nsims, birthday_problem(npeople = npeople)))/nsims
}
simulate_birthday_problem(nsims = 1000, npeople = 24)
```


```{r}
bday_probs <- sapply(seq(2, 100, 1), simulate_birthday_problem, nsims = 1000)
plot_data <- data.table(number_of_people_in_room = seq(2, 100, 1),
  probability_of_two_sharing_bday = bday_probs)
ggplot(plot_data, aes(x = number_of_people_in_room, 
  y = probability_of_two_sharing_bday)) +
  geom_point() + theme_bw() +
  ggtitle("Birthday Problem Simulations")
```


As our final trick, let's use simulations to show that OLS works. We'll build on the work we did above. First, we will write down the DGP for our data: it's normally distributed with some set parameter values (if this doesn't make perfect sense, don't worry! We'll come back to this often this semester). Second, we will draw a sample, and write down the coefficients. Third, we will do this a lot of times and show that OLS gives us the proper answers.

## OLS works!
Now let's verify that OLS gives us the right number in repeated samples!
```{r}
test_ols <- function(n, b0, b1, b2, seed = 
  sample(.Machine$integer.max, 1))
{
  x1 <- rnorm(n)
  x2 <- rnorm(n)
  y <- b0 + b1 * x1 + b2 * x2 + rnorm(n, 0, 1)
  coef(lm(y ~ x1 + x2))
}
test_ols(n = 100, b0 = 3, b1 = -2, b2 = 0.5)
simulations <- t(replicate(1000, test_ols(n = 1000, b0 = 3, b1 = -2, b2 = 0.5)))
plot_data <- melt(simulations)
real_answer_data <- data.frame(v1 = c(3, -2, 0.5), 
  Var2 = levels(plot_data$Var2))
ggplot(plot_data, aes(value)) + geom_histogram() + 
  facet_grid(~ Var2, scales = "free") +
  geom_vline(aes(xintercept = v1), data = real_answer_data) + theme_bw()

```

## Interactive Plot

```{r eruptions}
inputPanel(
  selectInput("n_breaks", label = "Number of bins:",
              choices = c(10, 20, 35, 50), selected = 20),
  
  sliderInput("bw_adjust", label = "Bandwidth adjustment:",
              min = 0.2, max = 2, value = 1, step = 0.2)
)

renderPlot({
  hist(faithful$eruptions, probability = TRUE, breaks = as.numeric(input$n_breaks),
       xlab = "Duration (minutes)", main = "Geyser eruption duration")
  
  dens <- density(faithful$eruptions, adjust = input$bw_adjust)
  lines(dens, col = "blue")
})
```


```


